### 2018-11-30 response/response-demography-1.tex

[I should have those somewhere already. I think they were in the original version.]

5. The use of competing hazard models is not new in demographic research
and doesn’t require long exposition. This section can be cut
sustainably. I don’t think the methodological innovation is a key
contribution of the paper as the methods are established even if they
have not been used to examine spacing.

[Response:]

I have reduced the length of the section by more than 1/4 (the original section was 5
pages long, while the revised section is less than 3.5).
There are three important differences between my approach and what is standard
in the literature, and I believe further shortening would negatively impact readers'
ability to understand the results.
First, most research that use survival models employ proportional hazard models, so 
motivating and providing more information on the non-proportional set-up is important.
Second, the use of conditional survival curves is, to my knowledge, new, and, therefore,
requires a detailed discussion.
Third, the sex ratio calculations is a new way to use competing risk estimates and is 
required to understand whether the use of sex selection vary within spells.

### 2018-11-30 paper/sex_selection_spacing-ver2.tex
Shorten Estimation Strategy section

\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}
\usepackage[title]{appendix}
\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{adjustbox}
\usepackage{xfrac}

@@

The unit of analysis is a spell, the period from one birth---or for the first spell from 
marriage---to the next birth.
The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The standard approach in the spacing literature is to use a single-exit model, but
access to sex selection means that the sex of the next child born is not random 
and that the spacing to girls and boys depends on parents' choices.
Both the non-randomness and differential spacing can be captured with a competing risk set 
up.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

I use a non-proportional hazard specification, which allows the shape of the hazard 
functions to vary across groups, rather than the standard proportional hazard function 
for three reasons.
First, it is unlikely---even in the absence of prenatal sex determination---that the 
effect of household characteristics such as the sex composition of previous births have 
the same effect throughout the entire spell.%
\footnote{
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
}
Second, the introduction of prenatal sex determination likely exacerbates any bias 
from the proportionality assumption since sex-selective abortion use varies across groups
and affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
Finally, the use of a non-proportional specification also mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is specified as
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

@@

Because the probability of ever having a next birth likely varies across groups, a direct 
comparison of standard survival curves tells us little about how the spread of sex selection 
affects birth spacing across groups.
I, therefore, condition on the predicted likelihood of parity progression when examining 
birth spacing measures, such as the median duration to a birth.
The reliability of this approach depends on whether the spell length covered is 
sufficiently long that few women are likely to give birth after the spell cut-off.
I discuss the choice of spell length below.%
\footnote{
It is important to note that the approach is not the same as merely calculating 
the birth spacing measures for women who already have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
In addition to calculating conditional birth spacing measures for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

Given the predicted probabilities of having a boy and of having a girl for each 
$t$ within a spell, is it easy to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.

>>>>

\usepackage{fontspec}
\setromanfont[Ligatures=TeX]{TeX Gyre Pagella}
\usepackage{unicode-math}
\setmathfont{TeX Gyre Pagella Math}
\usepackage{amsmath}
\usepackage[title]{appendix}
\usepackage[margin=1.0in]{geometry}
\usepackage[figuresleft]{rotating}
\usepackage[longnamesfirst]{natbib}
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[flushleft]{threeparttable}
\usepackage{setspace}
\usepackage[justification=centering]{caption}
\usepackage[font=scriptsize]{subfig}
\usepackage[xetex,colorlinks=true,linkcolor=black,citecolor=black,urlcolor=black]{hyperref}
\usepackage{adjustbox}
\usepackage{xfrac}

@@

The unit of analysis is a spell, the period from one birth---or for the first spell from 
marriage---to the next birth.
The model is a discrete time, non-proportional, competing risk hazard model with two exit 
states: either a boy or a girl is born.
The standard approach in the spacing literature is to use a single-exit model, but with
access to sex selection the sex of the next child is not random and parents' choices 
will impact the spacing to a girl or a boy differentially.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
Both the non-randomness and differential spacing can be captured with a competing risk set 
up.

I use a non-proportional hazard specification, which allows the shape of the hazard 
functions to vary across groups, rather than the standard proportional hazard function 
for three reasons.
First, it is unlikely---even in the absence of prenatal sex determination---that the 
effect of household characteristics such as the sex composition of previous births have 
the same effect throughout the entire spell.%
\footnote{
Assuming that the effect of sex composition is the same throughout a spell is especially 
problematic for higher-order spells where different sex composition of previous births can 
lead to substantial differences both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
}
Second, the introduction of prenatal sex determination likely exacerbates any bias 
from the proportionality assumption since sex-selective abortion use varies across groups,
which affects birth spacing, and because a household's use of sex selection may vary 
within a spell, which means that the effects of covariates vary within the spell as well.
Finally, the use of a non-proportional specification mitigates any potential effects 
of unobserved heterogeneity when used in conjunction with a flexible baseline hazard 
\citep{Dolton1995}.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is specified as
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where the explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, capture 
individual, household, and community characteristics discussed below,
and $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
with $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

@@

Because the probability of ever having a next birth likely varies across groups, a direct 
comparison of standard survival curves tells us little about how the spread of sex selection 
affects birth spacing across groups.
I, therefore, condition on the predicted likelihood of parity progression when examining 
birth spacing measures, such as the median duration to a birth.
The reliability of this approach depends on whether the spell length covered is 
sufficiently long that few women are likely to give birth after the spell cut-off.
I discuss the choice of spell length below.%
\footnote{
It is important to note that the approach is not the same as merely calculating 
the birth spacing measures for women who already have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
}
In addition, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Interpretation of the model coefficients is challenging \citep{thomas96}.
It is, however, possible to calculate the predicted probabilities of 
having a boy, $b$, and of having a girl, $g$, in period $t$, conditional on 
a set of explanatory variables and not having had a child before that period, as
\begin{align}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})}
\label{eq:probability_boy} \\
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
& =  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}
\label{eq:probability_girl}
\end{align}
It is then straightforward to calculate the estimated percentage of children born that 
are boys, $\hat{Y}$, at each $t$:  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
